### 2020-08-03 paper/sex_selection_spacing-ver4.tex
Edit Background and Conceptual Framework section

I divide eduction into four groups: no education, one to seven years, eight to eleven 
years, and twelve and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and eleven and twelve for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to sevens years of schooling completed and the eight to nine 
and ten to eleven years of schooling completed.
There is a trade-off between having groups large enough for the analyses below and having 
groups detailed enough to capture differences in behavior across education levels, such as 
the U-shaped relationship between education and labor force participation.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90 percent for the 1930s cohorts to 
less than 20 percent for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than sixty percent for the 1990s cohorts, with about half in the eight to eleven 
group and the other half in the twelve plus group.
In urban areas, the proportions with no education or one to seven years have declined 
to just below ten percent each.
Most of the increase in urban female education came from the twelve plus group, which 
now account for more than half of all urban women for the lastest cohorts.

@@

The standard economic argument for spacing becoming shorter with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

@@

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
better enables them to access prenatal sex determination and lowers the relative costs of 
using sex selection and have long birth intervals.

@@

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and non-existing for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is one crucial aspect of child quality---health and mortality---where
longer spacing does lead to better outcomes, although this relationship weakens with 
maternal education \citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, they adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation is an indication that this
process still operates.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better-educated 
because their household income increases the most---even with declining female labor 
participation---and their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition does not substantially change 
this group's use of sex selection.

>>>>

I divide eduction into four groups: no education, one to seven years, eight to eleven 
years, and twelve and more years.
The latter two correspond to having completed primary and completed secondary school, 
respectively.%
\footnote{
Although there are variations by state, elementary education in India consists of a 
primary school covering grades one through five and an upper primary---or middle 
school---covering grades six through eight. 
Similarly, secondary education covers grades nine and tenth for ``secondary education'' 
and eleven and twelve for ``upper secondary.''
}
To ensure that the results are comparable with the prior literature on fertility
and mortality in India, I follow the NFHS reports, except that I combine the 
less than five years and five to sevens years of schooling completed and the eight to nine 
and ten to eleven years of schooling completed.
This grouping allows me to capture the differences across education levels discussed below, 
while also having groups large enough for the empirical method.

% [what has happened to education]
Female education has increased substantially over time in rural and urban areas but 
is still substantially higher in urban than in rural areas.
Figure \ref{fig:education_over_time} shows the distribution of schooling by birth 
cohort for urban and rural women, 
twenty years or older, whether married or not, based on the four rounds of the NFHS.
In rural areas, women with no education have gone from 90 percent for the 1930s cohorts to 
less than 20 percent for the 1990s cohorts.
Women with eight or more years of education have gone from almost zero for the 1930 cohort 
to more than sixty percent for the 1990s cohorts, with about half in the eight to eleven 
group and the other half in the twelve plus group.
In urban areas, the proportions with no education or one to seven years have each declined 
to just below ten percent.
Most of the increase in urban female education came from the twelve plus group, which 
now account for more than half of all urban women for the lastest cohorts.

@@

The standard economic argument for shorter spacing with increasing female education 
is that parents incur time costs when they have children \citep{Hotz1997,schultz97}.
Specifically, if having children require the mother to reduce her market work, parents 
can lower the cost of children by shortening birth spacing to 
take advantage of economies of scale in childrearing \citep{Vijverberg1982}.

@@

The introduction of sex selection allows parents to avoid giving birth to girls but 
increases the expected duration to the next birth.
Theory suggests that sex selection increases with lower desired fertility and with
higher parity for a given desired number of children \citep{Portner2015b}.
Sex selection is more widespread among better-educated than less-educated and among urban 
than rural women, which is consistent with lower desired fertility increasing the
use of sex selection 
\citep{das_gupta97,retherford03b,Guilmoto2009a,Portner2015b,Jayachandran2017}.
Better-educated and urban women also tend to live in households with higher income, which 
lowers the relative costs of using sex selection and have long birth intervals.

@@

If more and ``better'' parental attention per child results in higher child ``quality,'' 
we should expect longer birth intervals \citep{Zajonc1975,Zajonc1976,Razin1980}.
However, the evidence on spacing's effect on child quality measures such as IQ 
and education is mixed for developed countries and non-existing for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
The exception is health and mortality, where longer spacing does lead to better outcomes, 
although this relationship weakens with maternal education 
\citep{Whitworth2002,Conde-Agudelo2012,Molitoris2019}.

@@

The increases in female educational attainment imply that access to education has 
expanded beyond the higher castes. 
One possible effect of the associated change in the composition of better-educated women 
is that this group's behavior would change.
However, ``Sanskritization'' implies that as lower castes females gain access to 
education and their husbands' income increases, they adopt higher caste 
norms such as stronger son preference and a retraction from the formal labor 
market \citep{Srinivas1956,Chen1995,Abraham2013,Chatterjee2018}.
The low and declining female labor force participation suggests that this process still operates.

@@

In summary, with substantial increases in husbands' income and a declining female labor 
force participation, I expect a push toward longer birth spacing over time, independent
of education levels, based on the income effects and the effects of spacing
on child outcomes.
Furthermore, I expect birth spacing to increase the most among the better-educated 
because their household income increases the most---even with declining female labor 
participation---and their use of sex selection.
Even with the substantial increase in the number of better-educated women, 
"Sanskritization" implies that the changing composition will not substantially change 
the use of sex selection.
